---
output: pdf_document
geometry: margin=1.5cm
header-includes:
  - \usepackage{booktabs}
  - \usepackage{tabu}
  - \usepackage{longtable}
  - \usepackage{caption}
---

\captionsetup[table]{labelformat=empty}

```{r setup-supp-tab, include = FALSE}
# knitr.table.format = "latex" not necessary
knitr::opts_chunk$set(
  cache.extra = knitr::rand_seed, 
  message = FALSE, 
  warning = FALSE, 
  error = FALSE, 
  echo = FALSE, 
  fig.pos = "H")
# disable sci number format
options(scipen = 999)
options(knitr.kable.NA = '')
# tools
library(here)
library(papaja)
library(captioner)
library(glue)
# data wrangling
library(dplyr)
library(tidyr)
library(tibble)
library(purrr)
# statistical modeling
library(lsr)
library(irr)
library(pROC)
library(caret)
library(xgboost)
# result plots
library(ggpubr)
library(ggplot2)
library(plotROC)
library(gtable)
# tables
library(knitr)
library(kableExtra)
# constant variables
emotion_list <- c(
  "anger", 
  "disgust", 
  "fear", 
  "happiness", 
  "sadness", 
  "surprise"
  )

classifier_level <- c(
  "Humans",
  "Affectiva",
  "CrowdEmotion",
  "Emotient", 
  "Microsoft",
  "MorphCast",
  "Neurodatalab",
  "VicarVision",
  "VisageTechnologies"
  )

name_convention_recode <- c(
  "Humans" = "Human Observers",
  "Affectiva" = "Affectiva",
  "CrowdEmotion" = "CrowdEmotion",
  "Emotient" = "Emotient",
  "Microsoft" = "Microsoft",
  "MorphCast" = "MorphCast",
  "Neurodatalab" = "Neurodata Lab",
  "VicarVision" = "VicarVision",
  "VisageTechnologies" = "Visage Technologies"
  )
```

S2 Table. Metrics for determining the confidence score (CS), the recognized emotion label, and the emotion classification score per video (B) based on the raw data (A) from the automatic classifiers.

**A**
```{r}
tibble::tribble(
    ~`$Video$`, ~`$Frame$`, ~`$Emotion Displayed$`, ~`$Emotion Recognized_1$`,    ~`$\\hdots$`, ~`$Emotion Recognized_j$`,
   "$Video_1$",    "$t_0$",       "$EmoDis_{1,1}$",  "$\\psi_{0,1,1}$",     "$\\hdots$",  "$\\psi_{0,1,j}$",
   "$Video_1$",    "$t_1$",       "$EmoDis_{1,1}$",  "$\\psi_{1,1,1}$",     "$\\hdots$",  "$\\psi_{1,1,j}$",
   "$Video_1$","$\\vdots$",            "$\\vdots$",               "$\\vdots$",     "$\\vdots$",               "$\\vdots$",
   "$Video_1$",    "$t_x$",       "$EmoDis_{1,1}$",  "$\\psi_{x,1,1}$",     "$\\hdots$",  "$\\psi_{x,1,j}$",
   "$Video_2$",    "$t_0$",       "$EmoDis_{2,2}$",  "$\\psi_{0,2,1}$",     "$\\hdots$",  "$\\psi_{0,2,j}$",
   "$Video_2$",    "$t_1$",       "$EmoDis_{2,2}$",  "$\\psi_{1,2,1}$",     "$\\hdots$",  "$\\psi_{1,2,j}$",
   "$Video_2$","$\\vdots$",            "$\\vdots$",               "$\\vdots$",     "$\\vdots$",               "$\\vdots$",
   "$Video_2$",    "$t_x$",       "$EmoDis_{2,2}$",  "$\\psi_{x,2,1}$",     "$\\hdots$",  "$\\psi_{x,2,j}$",
   "$\\vdots$","$\\vdots$",            "$\\vdots$",               "$\\vdots$",     "$\\vdots$",               "$\\vdots$",
   "$Video_i$",    "$t_0$",       "$EmoDis_{i,j}$",  "$\\psi_{0,i,1}$",     "$\\hdots$",  "$\\psi_{0,i,j}$",
   "$Video_i$",    "$t_1$",       "$EmoDis_{i,j}$",  "$\\psi_{1,i,1}$",     "$\\hdots$",  "$\\psi_{1,i,j}$",
   "$Video_i$","$\\vdots$",            "$\\vdots$",               "$\\vdots$",     "$\\vdots$",               "$\\vdots$",
   "$Video_i$",    "$t_x$",       "$EmoDis_{i,j}$",  "$\\psi_{x,i,1}$",     "$\\hdots$",  "$\\psi_{x,i,j}$"
  ) %>%
  knitr::kable(
    escape = FALSE, 
    booktabs = TRUE, 
    linesep = c('', '', '', '\\addlinespace', '', '', '', '\\addlinespace', '\\addlinespace', '', '', ''),
    align=rep('c', 6)
  ) %>%
  kableExtra::kable_styling(font_size = 7) %>% 
  kableExtra::column_spec(1, extra_css = "vertical-align:middle;")
```

**B**
```{r}
cs11 <- "$\\frac{\\sum_{x = 0}^{T}\\psi_{x,1,1}}{\\sum_{j = 1}^{J}\\sum_{x = 0}^{T}\\psi_{x,1,j}}$"
cs21 <- "$\\frac{\\sum_{x = 0}^{T}\\psi_{x,2,1}}{\\sum_{j = 1}^{J}\\sum_{x = 0}^{T}\\psi_{x,2,j}}$"
csi1 <- "$\\frac{\\sum_{x = 0}^{T}\\psi_{x,i,1}}{\\sum_{j = 1}^{J}\\sum_{x = 0}^{T}\\psi_{x,i,j}}$"
cs1j <- "$\\frac{\\sum_{x = 0}^{T}\\psi_{x,1,j}}{\\sum_{j = 1}^{J}\\sum_{x = 0}^{T}\\psi_{x,1,j}}$"
cs2j <- "$\\frac{\\sum_{x = 0}^{T}\\psi_{x,2,j}}{\\sum_{j = 1}^{J}\\sum_{x = 0}^{T}\\psi_{x,2,j}}$"
csij <- "$\\frac{\\sum_{x = 0}^{T}\\psi_{x,i,j}}{\\sum_{j = 1}^{J}\\sum_{x = 0}^{T}\\psi_{x,i,j}}$"

code1 <- "$\\left\\{\\begin{matrix} EmoDis_{1,1} = EmoRec_{1,j}, 1\\\\ EmoDis_{1,1} \\neq EmoRec_{1,j}, 0 \\end{matrix}\\right.$"
code2 <- "$\\left\\{\\begin{matrix} EmoDis_{2,2} = EmoRec_{2,j}, 1\\\\ EmoDis_{2,2} \\neq EmoRec_{2,j}, 0 \\end{matrix}\\right.$"
codei <- "$\\left\\{\\begin{matrix} EmoDis_{i,j} = EmoRec_{i,j}, 1\\\\ EmoDis_{i,j} \\neq EmoRec_{i,j}, 0 \\end{matrix}\\right.$"

tibble::tribble(
    ~`$Video$`,~`$Emotion Displayed$`, ~`$CS Emotion Recognized_1$`,    ~`$\\hdots$`, ~`$CS Emotion Recognized_j$`,   ~`$Emotion Recognized$`, ~`$Emotion Classification$`,
   "$Video_1$",      "$EmoDis_{1,1}$",                         cs11,     "$\\hdots$",                         cs1j,"$\\max(CS EmoRec_{1,j})$",                       code1, 
   "$Video_2$",      "$EmoDis_{2,2}$",                         cs21,     "$\\hdots$",                         cs2j,"$\\max(CS EmoRec_{2,j})$",                       code2, 
    "$\\vdots$",            "$\\vdots$",                   "$\\vdots$","$\\vdots$",                   "$\\vdots$",                "$\\vdots$",                  "$\\vdots$",
   "$Video_i$",      "$EmoDis_{i,j}$",                         csi1,     "$\\hdots$",                         csij,"$\\max(CS EmoRec_{i,j})$",                       codei
  ) %>% 
  knitr::kable(
    escape = FALSE, 
    booktabs = TRUE, 
    linesep = "",
    align=rep('c', 7)
  ) %>%
  kableExtra::kable_styling(font_size = 7)
```

Note: $i$ is one of the 937 videos processed, $x$ is a video frame at $t_x$, $j$ is an emotion label among *anger*, *disgust*, *fear*, *happiness*, *sadness* and *surprise*, and $\psi$ is the value of the odds ratio for the frame $t_x$ and for the emotion label $j$.